---
title: "Part and Partial Correlations in Regression"
author: "Christopher David Desjardins"
date: "2/5/2019"
output: html_document
---

The purpose of this activity is to explore the relationship between part and partial correlations and regression.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## R package for correlations
```{r}
install.packages("ppcor")
library(ppcor)
```


## Data
Holzinger-Swineford data set.

```{r}
hs.data <- read.csv("https://github.com/cddesja/epsy8266/raw/master/course_materials/data/HolzingerSwineford1939.csv")
```

## Activity

First, fit a regression by regressing deduct onto numeric. Then find the r-square

```{r}
mod.num <- lm(deduct ~ numeric, data = hs.data)
summary(mod.num)$r.square
```

Now, fit a regression by regressing deduct onto problemr and find the r squared. I have started the code

```{r}
mod.prob <- 

```

Then fit a multiple regression by regressing deduct onto numeric and problemr. Find the r squared.

```{r}
mod.mlr <- 
  
```

Question 1: What is the difference in r squared between `mod.num` and `mod.mlr` (Rsquared1) and between `mod.prob` and `mod.mlr` (Rsquared2)? 

Question 2: What is the proportion of variability not explained by `mod.num` (NotExp1) and `mod.prob` (NotExp2)?

Question 3: Calculate the ratio of the Rsquared1 to NotExp2 and the ratio of Rsquared2 to NotExp1.

Now, calculate the part and partial correlations between 
1. deduct and numeric paritalling out problemr
2. deduct and problemr paritalling out numeric
3. deduct and numeric partialling out problemr from numeric only
3. deduct and problemr partialling out numeric from problemr only

```{r}
# partial correlations
pcor(hs.data[ ,c("deduct", "numeric", "problemr")])$estimate
(pcor(hs.data[ ,c("deduct", "numeric", "problemr")])$estimate)^2


# part correlations
spcor(hs.data[ ,c("deduct", "numeric", "problemr")])$estimate
(spcor(hs.data[ ,c("deduct", "numeric", "problemr")])$estimate)^2
```

Question 4: What is the relationship between your answer to 1 and 3 and these correlations? How do you interpret these values?



