---
title: "Making lavaan estimate all the effects"
author: "Christopher Desjardins"
date: "Feburary 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In path analysis, we differentiate between several types of effects.

- __Direct effect__ - The influence of one variable on another that is unmediated by any other variable, i.e. each single headed arrow represents a direct effect
- __Indirect effects__ - Effect that is mediated by at least one intervening variable
- __Total causal effects__ - Sum of the above two effects
- __Spurious effects__ - Effects due to a common cause
- __Correlated effects__  - Effects that aren't analyzed in our model

If we add together the direct effect, indirect effects, spurious effects, and the correlated effects we are able to recover the bivariate correlation. We can make `lavaan` calculate them directly. By having `lavaan` do this as well, we could get (bootstrap) standard errors for these effects. 

We will just look at the model on slide #13 in the lecture notes (hopefully I don't change that number between now and then). 

First, we'll read in the data.

```{r}
slide13.dat <- read.table("https://raw.githubusercontent.com/cddesja/epsy8266/master/course_materials/data/decompose.dat", col.names = c(paste0("y", 1:3), paste0("x", 1:3)))
head(slide13.dat)
```

The correlation table on that slide is:

```{r}
cor(slide13.dat)
```

Let's use `lavaan` to estimate all these effects. We'll fit a model to calculate the correlation between `x1` and `y3` (0.508).

```{r}
library("lavaan")

effects.x1 <- '
y1 ~ a*x1 + b*x2
y3 ~ c*x1 + d*x2 + e*y1
x1 ~~ f*x2

# indirect effect of x1 through y1
ae := a*e

# spurious effects - none!

# correlated effects
df := d*f
ebf := e*b*f

# total causal effects
tce := c + (a*e)

# bivariate correlation
biv.cor.x := c + (a*e) + (d*f) + (e*b*f)
'

# std.ov standardizes all observed variables
effects.x1.fit <- sem(effects.x1, slide13.dat, std.ov = TRUE, se = "none")
summary(effects.x1.fit)
```

We see that we are able to recover our bivariate correlation.  

Now, see if you can recreate the correlation for

- `x2` and `y1` (0.572)
- `y1` and `y3` (0.847)





