---
title: "Motivation for logistic/probit regression"
author: "Christopher David Desjardins"
date: "2/7/2019"
output: html_document
---

Purpose: To motivate the use of logistic regression and/or probit regression.

# About Data
The data are from a Washington University twin data study look at the Minnesota Multiphasic Personality Inventory (MMPI) & schizophrenia. The data can be downloaded off the website.

The variables contained in the data are:

* Schizo = 'Consensus diagnosis'
* L   = 'Lie Scale'
* F   = 'F Scale'
* K   = 'K Scale'
* Hs  = 'Hypochondriasis'
* D   = 'Depression'
* Hy  = 'Hysteria'
* Pd  = 'Psychopathy'
* Mf  = 'Masc / Fem'
* Pa  = 'Paranoia'
* Pt  = 'Psychasthenia'
* Sc  = 'Schizophrenia'
* Ma  = 'Hypomania'
* Si  = 'Introversion'
* male = 1: male, 0: female

Information about these scales can be found here: https://psychcentral.com/lib/minnesota-multiphasic-personality-inventory-mmpi/.


The Schizo variable is coded as:

- 0 consensus diagnosis of not schizophrenia spectrum
- 1 consensus diagnosis of schizophrenia spectrum disorder
- 2 consensus diagnosis of definite or probable schizophrenia  

The Schizo variable is coded as:

- 0 consensus diagnosis of not schizophrenia spectrum
- 1 consensus diagnosis of definite or probable schizophrenia  

We will try and predict definite or probably schizophrenia vs. not schizophrenia spectrum  from the `Hs`  (Hypochondriasis) and `Ma` (Hypomania) scales and the a respondent's gender (`male`).

```{r}
wuschiz <- read.csv("https://raw.githubusercontent.com/cddesja/epsy8266/master/course_materials/data/wuschiz.csv")
```

1. Fit a multiple regression model predicting schizophrenia given the hypochondriasis, hypomania, and gender.

```{r}
mod <- lm(Schizo ~ Hs + Ma + male, data = wuschiz)
```

2. Visually assess the appropriateness of your model using the `plot()` function and answer the following questions: a) Should you do hypothesis testing? b) Do you have homogeneity of variance? c) Do you have normality of your residuals?

```{r}
plot(mod)
```

3. Obtain predicted values for your model and use the `summary()` function to look at the distribution of the predicted values. Do you see a problem?

```{r}
summary(predict(mod))
```
















Class questions

```{r}
mod.lr <- glm(Schizo ~ Hs + Ma + male, data = wuschiz, family = "binomial")
mod.pro <- glm(Schizo ~ Hs + Ma + male, data = wuschiz, family = "binomial"(link = "probit"))
summary(mod.lr)
summary(mod.pro)
summary(mod)
```
