---
title: "Motivation for logistic/probit regression"
author: "Christopher David Desjardins"
date: "2/7/2019"
output: html_document
---

Purpose: To motivate the use of logistic regression and/or probit regression.

# About Data
The data are from a Washington University twin data study of the Minnesota Multiphasic Personality Inventory (MMPI) & schizophrenia. The data can be downloaded off the website.

```{r include = TRUE}
wuschiz <- foreign::read.spss("wuschiz.sav", to.data.frame = TRUE)
```

The variables contained in the data are:

* Schizo = 'Consensus diagnosis'
* L   = 'Lie Scale'
* F   = 'F Scale'
* K   = 'K Scale'
* Hs  = 'Hypochondriasis'
* D   = 'Depression'
* Hy  = 'Hysteria'
* Pd  = 'Psychopathy'
* Mf  = 'Masc / Fem'
* Pa  = 'Paranoia'
* Pt  = 'Psychasthenia'
* Sc  = 'Schizophrenia'
* Ma  = 'Hypomania'
* Si  = 'Introversion'

Information about these scales can be found here: https://psychcentral.com/lib/minnesota-multiphasic-personality-inventory-mmpi/.


The Schizo variable is coded as:

- 0 consensus diagnosis of not schizophrenia spectrum
- 1 consensus diagnosis of schizophrenia spectrum disorder
- 2 consensus diagnosis of definite or probable schizophrenia  

We will try and predict definite or probably schizophrenia (2) vs. not schizophrenia spectrum (0) from the `Hs`  (Hypochondriasis) and `Ma` (Hypomania) scales and the `Si` (Social Introversion) dichotomized (just for illustration). First, we will remove the people that are on the spectrum, recode the `Schizo` variable, and do a median split for Si (`si.med`).

```{r}
wuschiz <- subset(wuschiz, Schizo != 1)
wuschiz$Schizo <- ifelse(wuschiz$Schizo == 2, 1, 0)
wuschiz$si.med <- ifelse(wuschiz$Si >= median(wuschiz$Si), 1, 0)
```

1. Fit a multiple regression model predicting schizophrenia given the hypochondriasis, hypomania, and dichotomized social introversion scales.

2. Visually assess the appropriateness of your model using the `plot()` function and answer the following questions: a) Should you do hypothesis testing? b) Do you have homogeneity of variance? c) Do you have normality of your residuals?

3. Obtain predicted values for your model and use the `summary()` function to look at the distribution of the predicted values. Do you see a problem?


